\begin{thebibliography}{123}

\bibitem{AshteBL02}
A. Ashtekar, C. Beetle, and J.~Lewandowski : {\em Geometry of
generic isolated horizons},
Class. Quantum Grav. {\bf 19}, 1195 (2002).

\bibitem{AshteFK00}
A. Ashtekar, S. Fairhurst, and B. Krishnan : {\em Isolated horizons:
Hamiltonian evolution and the first law},
Phys. Rev. D {\bf 62}, 104025 (2000).

\bibitem{AshteK04}
A. Ashtekar and B. Krishnan : {\em Isolated and dynamical horizons
and their applications},
Living Rev. Relativity {\bf 7}, 10 (2004) \\
\url{http://www.livingreviews.org/lrr-2004-10}

\bibitem{Bard15}
G.V. Bard : {\em Sage for Undergraduates}, Americ. Math. Soc. (2015); \\
preprint freely downloadable from \url{http://www.gregorybard.com/}

\bibitem{Berge03}
M. Berger : {\em A Panoramic View of Riemannian Geometry},
Springer, Berlin (2003).

\bibitem{Carro04}
S.M. Carroll : \emph{Spacetime and Geometry: An Introduction to General Relativity},
Addison Wesley (Pearson Education), San Fransisco (2004); \\
\url{http://preposterousuniverse.com/spacetimeandgeometry/}

\bibitem{Carte69}
B.~Carter : {\em Killing horizons and orthogonally transitive groups
in space-time},
J. Math. Phys. {\bf 10}, 70 (1969).

\bibitem{Carte73}
B.~Carter : {\em Black hole equilibrium states Part II. General theory of stationary black hole states}, in {\em Black Holes -- Les astres occlus},  edited by C.~deWitt and B.~deWitt,
Gordon and Breach, New York (1973), p.~125; reprinted in
Gen. Relativ. Gravit. {\bf 42}, 653 (2010)\\
DOI: \href{http://dx.doi.org/doi:10.1007/s10714-009-0920-9}{10.1007/s10714-009-0920-9}

\bibitem{Carte87}
B.~Carter : {\em Mathematical foundations of the theory of
relativistic stellar and black hole configurations},
in {\em Gravitation in Astrophysics}, Eds. B.~Carter and J.B.~Hartle,
Plenum Press, New York (1987).

\bibitem{Carte97}
B. Carter : {\em Extended tensorial curvature analysis for embeddings
and foliations},
Contemp. Math. {\bf 203}, 207 (1997).

\bibitem{Choqu09}
Y. Choquet-Bruhat : {\em General Relativity and Einstein's Equations},
Oxford Univ. Press, New York (2009).

\bibitem{Choqu15}
Y. Choquet-Bruhat : {\em Introduction to General Relativity, Black Holes, and
Cosmology}, Oxford Univ. Press, New York (2015).

\bibitem{ChoquDD77}
Y. Choquet-Bruhat, C. De Witt-Moretten, and M. Dillard-Bleick :
{\em Analysis, Manifolds and Physics},
North-Holland, Amsterdam (1977).

\bibitem{Chrus02}
P.T. Chru\'sciel : {\em Black holes},
in {\em Proceedings of the T\"ubingen conference on conformal structure of
space-time}, edited by J.~Frauendiener, H.~Friedrich,
Lecture Notes in Physics {\bf 604}, 61,
Springer, Berlin (2002).

\bibitem{Chrus05}
P.T. Chru\'sciel : {\em Black holes -- An Introduction},
in {\em 100 Years of Relativity}, edited by A. Ashtekar,
World Scientific, Singapore (2005), p. 93.\\
DOI: \href{http://dx.doi.org/doi:10.1142/9789812700988_0004}{10.1142/9789812700988\_0004}

\bibitem{ChrusLH12}
P.T. Chru\'sciel, J. Lopes Costa, and M. Heusler :
{\em Stationary Black Holes: Uniqueness and Beyond},
Living Rev. Relativity {\bf 15}, 7 (2012) \\
\url{http://www.livingreviews.org/lrr-2012-7}

\bibitem{DaferR13}
M. Dafermos and I. Rodnianski : {\em Lecture on black holes and linear waves},
in  Clay Math. Proc. {\bf 17}: {\em Evolution Equations}, edited by
D. Ellwood, I. Rodnianski, G. Staffilani and J. Wunsch, AMS / Clay Mathematical Institute,
Cambridge (2013) \\
\url{http://people.maths.ox.ac.uk/cmi/library/proceedings/cmip017c.pdf}

\bibitem{Damou79}
T. Damour : {\em Quelques propri\'et\'es m\'ecaniques, \'electromagn\'etiques,
thermo\-dy\-na\-mi\-ques et quantiques des trous noirs},
Th\`ese de doctorat d'\'Etat, Universit\'e Paris 6 (1979); available at\\
\url{http://pagesperso.ihes.fr/~damour/Articles/}

\bibitem{Damou82}
T. Damour : {\em Surface effects in black hole physics},
in {\em Proceedings of the Second Marcel Grossmann Meeting on General
Relativity}, Ed. R.~Ruffini, North Holland (1982), p.~587; available at\\
\url{http://pagesperso.ihes.fr/~damour/Articles/}

\bibitem{Damou04}
T. Damour : {\em The Entropy of Black Holes: A Primer},
in {\em Poincaré Seminar 2003: Bose-Einstein Condensation -- Entropy},
edited by J. Dalibard, B. Duplantier and V. Rivasseau, Birkha\"auser, Basel (2004),
p.~227

\bibitem{Derue09}
N. Deruelle : {\em Black Holes in General Relativity}, lectures at
Institut de Physique Théorique, CEA, Saclay (France) (2009); slides and notes
available at \\
\url{http://ipht.cea.fr/Docspht/search/article.php?id=t09/346}

\bibitem{DerueU14}
N. Deruelle and J.-P. Uzan: {\em Th\'eories de la Relativit\'e}
Belin, Paris (2014).

\bibitem{Eschr11}
H. Eschrig : {\em Topology and Geometry for Physics},
Springer, Berlin (2011).

\bibitem{FroloZ11}
V.P. Frolov and A. Zelnikov : {\em Introduction to Black Hole Physics},
Oxford University Press, New York (2011).

\bibitem{GerocH82}
R. Geroch and J.B. Hartle : {\em Distorted black holes},
J. Math. Phys. {\bf 23}, 680 (1982).

\bibitem{Gourg13}
E. Gourgoulhon : {\em Special Relativity in General Frames; From Particles to Astrophysics},
Springer, Berlin (2013) \\
\url{http://relativite.obspm.fr/sperel}

\bibitem{Gourg14}
E. Gourgoulhon : {\em Relativit\'e g\'en\'erale},
Master Lecture Notes\\
\url{http://luth.obspm.fr/~luthier/gourgoulhon/fr/master/relat.html}

\bibitem{GourgJ06}
E. Gourgoulhon and J.L. Jaramillo : {\em A 3+1 perspective on null hypersurfaces and isolated horizons},
Phys. Rep. {\bf 423}, 159 (2006).

\bibitem{GourgJ08}
E. Gourgoulhon and J. L. Jaramillo : {\em New theoretical approaches to black holes},
in Proceedings of the Conference {\em Jean-Pierre Lasota, X-ray Binaries, Accretion Disks and Compact Stars} held in Trzebieszowice, Poland (7-13 October 2007), edited by M. Abramowicz and O. Straub, New Astronomy Reviews {\bf 51}, 791 (2008) \\
DOI: \href{http://dx.doi.org/doi:10.1016/j.newar.2008.03.026}{10.1016/j.newar.2008.03.026}

\bibitem{Hajic73}
P. H\'a\'\j i\v{c}ek : {\em Exact models of charged black holes. I. Geometry
of totally geodesic null hypersurface},
Commun. Math. Phys. {\bf 34}, 37 (1973). \\
DOI: \href{http://dx.doi.org/10.1007/BF01646542}{10.1007/BF01646542}

\bibitem{Hajic74}
P. H\'a\'\j i\v{c}ek : {\em Can outside fields destroy black holes ?},
J. Math. Phys. {\bf 15}, 1554 (1974). \\
DOI: \href{http://dx.doi.org/10.1063/1.1666846}{10.1063/1.1666846}

\bibitem{Hartl03}
J.B. Hartle : \emph{Gravity: An Introduction to Einstein's General Relativity},
Addison Wesley (Pearson Education), San Fransisco (2003); \\
\url{http://wps.aw.com/aw_hartle_gravity_1/}

\bibitem{Hawki71}
S.W.~Hawking : {\em Gravitational radiation from colliding black holes},
Phys. Rev. Lett. {\bf 26}, 1344 (1971).

\bibitem{Hawki72}
S.W.~Hawking : {\em Black holes in general relativity},
Commun. Math. Phys. {\bf 25}, 152 (1973).

\bibitem{Hawki73}
S.W.~Hawking : {\em The event horizon}, in
{\em Black holes, les astres occlus}, Eds. C.~DeWitt
and B.S.~DeWitt, Gordon and Breach, New York (1973).

\bibitem{HawkiE73}
S.W.~Hawking \& G.F.R. Ellis : {\em The large scale structure of
space-time},
Cambridge University Press, Cambridge (1973).

\bibitem{Hawki94}
S.W.~Hawking : {\em The Nature of Space and Time} (1994)\\
\url{http://arxiv.org/abs/hep-th/9409195}

\bibitem{HawkiP15}
S.W.~Hawking and R. Penrose: {\em The Nature of Space and Time},
Princeton Univ. Press, Princeton (2015).

\bibitem{Jaram13}
J.L. Jaramillo : {\em An Introduction to Local Black Hole Horizons in the 3+1
Approach of General Relativity}, in {\em Black Holes: New Horizon}, edited
by S.A. Hayward, World Scientific, Singapore (2013).

\bibitem{JoyneS14}
D. Joyner and W. Stein : {\em Sage Tutorial}, CreateSpace (2014).

\bibitem{Lafon15}
J. Lafontaine : {\em An Introduction to Differential Manifolds},
Springer, Cham (2015).

\bibitem{Langl13}
D. Langlois : \emph{Relativit\'e g\'en\'erale},
Vuibert, Paris (2013).

\bibitem{Lee97}
J. M. Lee : {\em Riemannian Manifolds: An Introduction to Curvature},
Springer, New-York (1997).

\bibitem{Lee13}
J. M. Lee : {\em Introduction to Smooth Manifolds}, 2nd edition,
Springer, New-York (2013).

\bibitem{Milno56}
J. W. Milnor : {\em On manifolds homeomorphic to the 7-sphere},
Ann. Math. {\bf 64}, 399 (1956).

\bibitem{MisneTW73}
C.W.~Misner, K.S.~Thorne, and J.A.~Wheeler : {\em Gravitation},
Freeman, New York (1973).

\bibitem{ONeil83}
B. O'Neill : {\em Semi-Riemannian Geometry, with Applications to Relativity},
Academic Press, New York (1983).

\bibitem{ONeil95}
B. O'Neill : {\em The geometry of Kerr black holes}, A.K. Peters, ?? (1995);
reprinted by Dover (2014).

\bibitem{Papan09}
E. Papantonopoulos (ed.) : {\em Physics of Black Holes}, Springer, Berlin (2009)

\bibitem{Penro65}
R. Penrose : {\em Gravitational collapse and space-time singularities},
Phys. Rev. Lett. {\bf 14}, 57 (1965).

\bibitem{Poiss04}
E. Poisson : \emph{A Relativist's Toolkit,
The Mathematics of Black-Hole Mechanics},
Cambridge University Press, Cambridge (2004); \\
\url{http://www.physics.uoguelph.ca/poisson/toolkit/}

\bibitem{SteinJ05}
W. Stein and D. Joyner : {\em SAGE: System for Algebra and Geometry Experimentation},
Commun. Comput. Algebra {\bf 39}, 61 (2005).

\bibitem{Strau04}
N. Straumann : \emph{General Relavity, with Applications to Astrophysics},
Springer-Verlag, Berlin (2004).

\bibitem{Taube87}
C. H. Taubes : {\em Gauge theory on asymptotically periodic 4-manifolds},
J. Differential Geom. {\bf 25}, 363 (1987).

\bibitem{Thorn94}
K.S. Thorne : {\em Black Holes and Time Warps}, Norton (1994).

\bibitem{Towns97}
P.K. Townsend : {\em Black Holes}, Lecture notes for a course taught in Part III
of the Cambridge University Mathematical Tripos (1997) \\
\url{http://arxiv.org/abs/gr-qc/9707012}

\bibitem{Wald84}
R.M.~Wald : {\em General relativity},
University of Chicago Press, Chicago (1984).

\bibitem{Wald01}
R.M. Wald :{\em The thermodynamics of black holes},
Living Rev. Relativity {\bf 4}, 6 (2001) \\
\url{http://www.livingreviews.org/lrr-2001-6}

\bibitem{Zimme13}
P. Zimmermann et al. : {\em Calcul math\'ematique avec Sage}, CreateSpace (2013) \\
freely downloadable from \url{http://sagebook.gforge.inria.fr/}

\end{thebibliography}
